---
title: "Do Economic Factors Affect Stock Prices"
author: "Raphael Tam"
date: "December 29, 2014"
output: html_document
---
# Analysis

```{r packages}
library(reshape2)
library(gridExtra)
library(memisc)
library(ggplot2)
library(ggthemes)
library(scales)
library(zoo)
library(GGally)
```
###  Load data from my file.  
Please refer to "get_data.Rmd" on how I constructed the dataset.
```{r load_the_data}

df<-read.csv("my_data.csv", header=TRUE)
head(df,3)
tail(df,3)
```
### Summary of data
The data is obtained from Quandl.com. It contains quarterly US macro-economic data from Q1 1969 to Q3 2014 and.  The sp_500 is a weighted index of the stock prices of the largest 500 companies in the US.  As an index, it has no unit.  Gross domestic product (gdp) is in billion dollars.  gdp_growth_rate' and 'sp_500_qq_change' are quarterly change rates.  'cum_inflation' is also an index.  The values in the gdp column can be divided by cum_inflation to get GDP in constant dollars.  'housing_start' is the number in thousands of permits granted nationally for that quarter. The unemployment rate (unempl_rate) is the % of work force that is unemployed and actively seeks jobs. 'president','senate' and 'house' are categorical variables.  The value is 1 when Democrats control one of these 3 branches, 0 otherwise.  Data summarized below.

```{r summary_of_data}
str(df)

```
## Single Variable 
Plot S&P500 over time first since this is the variable of interest
```{r single_variable}
#theme_set(theme_minimal(24))
ggplot(data=df,aes(x=as.Date(Date), y=sp_500))+
  geom_line(color="red")
###
```
Start by learning the range and distribution of single variables.  These variables are quarter to quarterly change rates: sp500, gdp growth,unemployment and housing start. Will look at whether these quarterly changes have an impact on the quarterly changes in the stock market as the second step.

Look at the SP500 quarterly channges first.  It looks remarkably symmetrical about 0. 
```{r sp500 quarterly change}
summary(df$sp500_qq_change)
ggplot(df, aes(x=sp500_qq_change))+
  geom_histogram(binwidth=.01, color="black",fill="white")  
###
```
GDP quarterly growth looks normally distributed.
```{r gdp growth rate}

summary(df$gdp_growth_rate)
ggplot(df, aes(x=gdp_growth_rate))+ 
  geom_histogram(binwidth=.0025,color='black',fill='white')
###
```
Unemployment is more broadly distributed than the GDP and S&P500
```{r unemployment_rate}
summary(df$unempl_rate)
ggplot(df, aes(x=unempl_rate)) + 
  geom_histogram(binwidth = .25, color="black",fill='white')
###
```
Housing starts also looks broader than GDP and the SP.  
```{r housing_starts}
summary(df$housing_start)
ggplot(df, aes(x=housing_start)) + 
  geom_histogram(binwidth = 100, color ='black',fill='white')
###
```
Take a look at cum_inflation to make sure it is monotonic.  It looks right.
```{r}
ggplot(data=df, aes(x=as.Date(Date), y=cum_inflation))+
  geom_line()
###
```
Now get quarter on quarter change in inflation since I hope to see clues of relationships among quarterly changes.
```{r}
#'getrate' was used in 'get_data.Rmd' to calculate the quarterly change rates for the sp_500 and gdp.  Reuse it here to get the quarterly change in inflation rate.

getrate <- function (ts){
  ts1<-c(ts,0)
  ts2<-c(0,ts)
  ts1<-ts1[2:(length(ts1)-1)]
  ts2<-ts2[2:(length(ts2)-1)]
  (ts2-ts1)/ts1
  }

q_inf_rate<-getrate(df$cum_inflation)

#set the last data point to 0 since it is the baseline
q_inf_rate[nrow(df)]=0

summary(q_inf_rate)

#Bind quarterly inflation rate to the data frame
df<-cbind(df, q_inf_rate)
```
Plot the histogram of quarterly inflation rate to get a feel for its distribution.
```{r}
ggplot(df, aes(x=q_inf_rate))+
  geom_histogram(binwidth=.001, color='black',fill='white')
####
```
Visualize below the quarterly change in inflation rate in time. Quarter to Quarter changes should not be monotonic.  Indeed, there are periods of deflation (negatvie inflation) but they are rare.  The one seen in 2009 was caused by the bank crisis.
```{r visualize quarterly changes}

ggplot(df, aes(x=as.Date(Date),y=q_inf_rate))+geom_line()
###
```
Now visualize party control of the 3 branches.  The plots below simply shows Democractic control (shown as high) versus Republican (low) over time. 
``` {r party control}

p1<-ggplot(data=df,aes(y=house,x=as.Date(Date)))+
  geom_step()+
  labs(title="Democratic control in House")+
  ylab("control=high")+
  scale_y_discrete(breaks=NULL)

p2<-ggplot(data=df,aes(y=senate,x=as.Date(Date)))+
  geom_step()+
  labs(title="Democratic control in Senate") +
  ylab("control=high")+
  scale_y_discrete(breaks=NULL)

p3<-ggplot(df)+
  geom_step(aes(y=df$president,x=as.Date(df$Date)))+
  labs(title="Democrat in the WH")+
  ylab("control = high")+
  scale_y_discrete(breaks=NULL)

grid.arrange(p1,p2,p3, ncol=1)
###
```
Do the same with Republicans control.  The high and lows should be reversed from the plots above.
```{r Republican control}
p1<-ggplot(df)+geom_step(aes(y=((df$house+1)%%2),
                             x=as.Date(df$Date)))+
  labs(title="Republican control in House") +
  ylab("control=high")+
  scale_y_discrete(breaks=NULL)

p2<-ggplot(df)+geom_step(aes(y=((df$senate+1)%%2),
                             x=as.Date(df$Date)))+
  labs(title="Republican control in Senate")+
  ylab("control=high")+
  scale_y_discrete(breaks=NULL)

p3<-ggplot(df)+geom_step(aes(y=((df$president+1)%%2),
                             x=as.Date(df$Date)))+
  labs(title="Republican in the WH")+
  ylab("control=high")+
  scale_y_discrete(breaks=NULL)

grid.arrange(p1,p2,p3, ncol=1)
  
####
```
Get a summary with tables. Democratic control is a 1.  So Democrats control the House and Senate more often than the Republicans, but less so for the White House 
```{r party control of 3 branches}
table (df$house)
table (df$senate)
table (df$president)
```
##Explore relationships between 2 variables 
```{r two_variables}
#Normalize the values to get a unified and compressed scale.
df[c('n_sp_500',
     'n_sp500_qq_change',
     'n_gdp',
     'n_gdp_growth_rate',
     'n_housing_start', 
     'n_unempl_rate',
     'n_q_inf_rate')] = 
  scale(df[c('sp_500',
             'sp500_qq_change',
             'gdp',
             'gdp_growth_rate',
             'housing_start',
             'unempl_rate',
             'q_inf_rate')],
              center=T,scale=T)
######
```
Does the gdp growth rate "explain" quarterly changes in the S&P 500?
See the sp500 quarter on quarter change against quarterly gdp rate below.
```{r melt and color}

#First melt data into long frame in order to color code the line plots

df_long = melt(subset(df, select =c(Date,
                                    n_sp500_qq_change, 
                                    n_gdp_growth_rate)),
               id.vars="Date",
               variable.name="rates",
               value.name="rate_values")
```
Check below for temporal relationship between the two rates
```{r temporal relationship between quarterly S&P and GDP}
ggplot(data=df_long, aes(x=as.Date(Date),
                         y=rate_values,
                         color= rates)) +
  geom_line(aes(group=rates))
####
```
Don't see any pattern in the plot above.  How about inflation rate?
```{r more 2 variable plots}

ggplot(data=df, aes(x=n_sp500_qq_change,
                    y=n_q_inf_rate)) +
                geom_point()

cor.test(df$gdp_growth_rate, df$q_inf_rate)
#####
```
No discernable relationship between the gdp growth and inflation rates either.  How about the relationship between governing party and the stock index?  The calculation and the plot below suggest that the SP500 has a higher mean and fewer negative outliers with Democratic Presidents.  
```{r boxplots by governing party}
by(df$sp500_qq_change, df$president, summary)
ggplot(df, aes(x=as.factor(df$president), 
               y=df$sp500_qq_change), 
               fill=df$president)+
      geom_boxplot()
####
```
Do the same with House and Senate control below. I can see from the plots below that the picture is reversed with Democratic control of House and Senate.  So it is difficult to draw any inference on the effect of the governing party on quarterly changes in the sp500. 
``` {r House and Senate control}

ggplot(df, aes(x=as.factor(df$house), 
               y=df$sp500_qq_change), 
       fill=as.factor(df$house))+geom_boxplot()
ggplot(df, aes(x=as.factor(df$senate), 
               y=df$sp500_qq_change), 
       fill=df$senate)+
       geom_boxplot()

###
```
Seems to go into deadends with the governing philosophy and econoimic variables. Since the stock market may anticpate economic growth, does the sp500 quarter on quarter change reflect the values of economic variables 1 or two quarters earlier?  
Let me check the pearson correlation between the S&P and the values of GDP from the current as well as 1 and 2 previous quarters.
```{r delayed response from gdp growth}
#function for checking correlations delayed by 1 or 2 quarters.  The function will check nq-1 number of quarters back.
sp5_var_cor<-function(df, ts, nq){
  for (i in 1:nq){
    j=i-1
    rate_advanced<-ts[(j+1):(j+nrow(df))]
    df = cbind(df,rate_advanced)
    print (cor.test(df$sp500_qq_change,rate_advanced))
  }
}
sp5_var_cor(df,df$gdp_growth_rate,3)
```
Don't see a pattern between the S&P and the GDP quarterly change rates from the correlation coefficients above.  How about other econoomic variables, namely, quarterly inflation rate, housing start and unemployment rate?  Check the correlation and also plot sp_500 change against these variables for the same quarter.

First check quarterly inflation
```{r quarterly inflation}
sp5_var_cor(df, df$q_inf_rate, 3)

ggplot(data=df, aes(x=q_inf_rate, y=sp500_qq_change))+
geom_point()
###
```
No pattern with quarterly inflation rate seen in the plot above. How about housing start?
```{r housing_start}
sp5_var_cor(df, df$housing_start, 3)

ggplot(data=df, aes(x=housing_start, y=sp500_qq_change))+
geom_point()
###
```
No pattern with housing start.  How about umemployment rate?
```{r unemployment}
sp5_var_cor(df, df$unempl_rate, 3)

ggplot(data=df, aes(x=unempl_rate, y=sp500_qq_change))+
geom_point()
####
```
I haven't seen anything intersting so far.  These quarterly economic changes show little correlation with sp_500 quarterly variations for the current quarter or up to 2 quarters back.
The time lag theory doesn't seem to produce clues on how the S&P quaterly changes are related to economic variables.  
I wonder what the scatter plot between the sp500 index itself (not the quarterly change) and the gdp look like?  

```{r gdp and sp_500 plot}
ggplot(data=df, aes(x=gdp,y=sp_500))+
  geom_point()

with (df, cor.test(sp_500, gdp))
```
Bingo! The plot and calculation above shows that the GDP is highly correlated to the SP500 itself, although it is a complex relationship.  There is little dispersion in the earlier periods.  It looks similar to the time series itself.  

Now let me see if there is a pattern between the various variables and the sp_500 itself.  Let me also color code the data with time to better discern the effect of time. 


```{r time marking}
#Period_index divides 47 years of history by 10 years except for the first 7 years.  Since data is in reverse chronological order, the most recent ten years (Q3/2014 to Q4/2003) is period 1.  Period 5 is from Q3/1973 to Q1/1967.  So except for period 5, each color codes 10 years worth of data.
df["period_index"]=c(rep(1,40),rep(2,40),rep(3,40),
                     rep(4,40),rep(5,31))
```
First check the S&P against inflation.  The data is color coded differently every 10 years.
```{r inflation rate against sp500 itself}
ggplot(data=df, aes(x=q_inf_rate, y=sp_500, 
                    color=period_index))+
  geom_point() +
  scale_y_continuous(trans=log10_trans())
####
```
That looks interesting.  The negative correlation is visible when it is below 1%.  Can I get a better picture by looking only at the data points below 1%?  
```{r zoom in lower inflation data}
ggplot(data=df, aes(x=q_inf_rate, y=sp_500, 
                    color=period_index))+
  geom_point()+
  scale_y_continuous(trans=log10_trans())+
  coord_cartesian(xlim=c(0, .01))
####
```
The plot above seems to show that the stock market likes lower inflation rate, but the dispersion is large.  The data points seem to concentrate in period 1, 2 and 3.  Let me look at the inflation data over time.
```{r inflation rate over time}
ggplot(data=df, aes(x=as.Date(Date), y=q_inf_rate))+geom_point()
####
```
The plot above suggests that inflation rate itself seems to decrease over time.  The Fed seems to be more effective in controlling inflation (and no deflation) in recent history and it may have an effect on the S&P.  I didn't know about that before.

Now look at Housing Starts against the S&P.  
```{r housing start}
ggplot(data=df, aes(x=housing_start, y=sp_500, color=period_index))+
geom_point()
####
```
The Housing Starts plot above suggests that it is positively correlated to the sp_500 but in different clusters.  I can see the historically low housing start in the last 6 years due to the real estate bubble collapse.  

Now look at unemployment.  
```{r unemployment rate}
ggplot(data=df, aes(x=unempl_rate, y=sp_500,color=period_index))+
geom_point()

####
```
The plot above clearly shows the negative correlation of unemployment with the sp500 in two clusters: one in period 1 and 2, with another one in period 2 and 3.  Unemployment seems to be uncorrelated with the stock market in periods 4 and 5.  So it apperas that like other economic factors, unemployment is related to the S&P in clusters.

Where are we?  The scatter plots of economic factors and the sp500 consistently show patterns in clusters, while the S&P itself keeps growing.  The 2 variable scatter plots colored with time suggest that there are time spans when some of these variables could be correlated to each other.  

What is remarkable is the rise of the SP over the long arc of 47 years. So before going further, I want to guess what could be affecting the S&P?  Is it possible that the economic variables affect quarterly changes in the short-term, but over the lonng term, something else more fundamental is driving stock prices?  The one thing that grows nonstop over the last 47 years is population.  A larger population should produce a higher GDP which in turn drives the S&P as shown earlier.  Let me get population data and see how sp_500 changes with the population.  
```{r get population data}
#US population data is availabe from the Census Bureau as a text document.  I downloaded the text data into 'us_pop.txt' from the Census Bureau.  I needed to sum the data by zip code by state by year.  I used python to do the extraction and wrote the total population by year into 'us_population1.csv'.  See 'load_population.py' on how I assemble population data.

pop = read.csv('us_population1.csv',header=T)

#Population data is available only yearly from 1969 to 2012 in chronological order.  So I have to wrangle the data in order to align it with the working data frame, which contains quarterly data and is in reverse chronlogical order.
pop_end_yr = 1969
pop_str_yr = 2012
df_end_yr = 1967
df_str_yr = 2014
str_row = (df_str_yr - pop_str_yr) * 4
end_row = nrow(df) - (pop_end_yr - df_end_yr) * 4
df_n<-subset(df,select=c(Date,sp_500,gdp))
df_s<-df_n[str_row:end_row,]
sp_500=df_s['sp_500']
n_quarters = (pop_str_yr - pop_end_yr + 1) * 4
sp_500_yr=sp_500[seq(1,n_quarters,4),]
gdp_q=df_s['gdp']
gdp_yr=gdp_q[seq(1,n_quarters,4),]
pop['sp500']<-rev(sp_500_yr)
pop['gdp']<-rev(gdp_yr)
#Now I can scatter plot population against sp_500 and gdp data from df, the running data frame.
ggplot(data=pop,aes(x=pop,y=gdp))+geom_point()
###
```
The plot above confirms that the gdp and the populaton are highly correlated.
```{r}
ggplot(data=pop, aes(x=pop,y=sp500))+geom_point()
###
```
The plot above shows the stock market and the population are highly correlated also.

The GDP bears an almost linear relationship with population growth.  An earlier plot shows the sp500 is highly correlated to the gdp. So it seems intuitive that as the popluation grows, the nation produces a higher gdp, whic in turns has a positive impact on the sp500.  Now that we have a better idea of what is positively correlated with the S&P, let me return to df.

Since the economic variables are clustered by time periods, what if I segregate the data by periods? I may see how the SP500 is impacted by the economic variables WITHIN the cluster.

The period_index is rather arbitrary, I would use the changes in party control of the White House as time markers instead.  It might provide some historical context since it is easier for me to think of the "Clinton years" rather than 1992 to 2000.  I will mark time periods by the party in the White House (potus.buckets).
```{r mark time by executive branch governing party}
df['potus.buckets']<-cut(as.numeric(
  format(as.yearmon(as.Date(df$Date)),"%Y")),
  breaks = c(1966,1969,1977,1981,1993,2001,2009,2015))
```
Now visualize how the S&P is related to the quarterly inflation rate, housing starts and unemployment.  These are essentially the same as the earlier plots, only color coded now with different period boundaries.

```{r}
ggplot(data=df, aes(x=q_inf_rate, y=sp_500,
                    color=potus.buckets, type="div"))+
  geom_point()

ggplot(data=df, aes(x=housing_start,y=sp_500,
                    color=potus.buckets,type="div"))+
  geom_point()

ggplot(data=df, aes(x=unempl_rate, y=sp_500, 
                    color = potus.buckets, type="div"))+
  geom_point()
####
```
The 3 plots above confirm that the relationships are clustered by time periods, and they do not have the same shape across all clusters.  The fact that there are many clusters and shapes explain why cor.tests show little correlation, as confirmed by the calculations below.
```{r correlation against sp_500}
with (df, cor.test(sp_500, q_inf_rate))
with (df, cor.test(sp_500, housing_start))
with (df, cor.test(sp_500, unempl_rate))
```
## 3 or more variables
Up to now, I have been visualizing the s&p against one economic variable over time in one scatter plot . How do I visualize the effect of 1 more variable?  I will use the size of the dot to visualize an additional variable.  The plot below shows GDP, inflation against the S&P, color coded by the potus time marker.
```{r three_variables}

ggplot(data=df, aes(x=gdp_growth_rate, y=sp_500, 
                    color=potus.buckets,type='div'))+
  geom_point(aes(size=q_inf_rate+5))+
  scale_y_continuous(trans=log10_trans())+
  geom_hline(yintercept=150, color="red")
#####
```
The plot above suggests that there may be 2 S&P growth mechanisms since the data appears as clusters below and above the red line. What is interesting is that inflation is consistently lower above the red line.

I will try another method to visualize unemployment rate, sp_500, qarterly inflation rate and time by using facet_wrap instead of color to segregate data by period. This will help me choose the visualization tool for the final plots.
```{r 3 variables}
ggplot(data=df, aes(x=n_unempl_rate,y=sp_500)) +
  scale_y_continuous(trans=log10_trans()) +
  geom_point(aes(size=n_q_inf_rate+5))+
  facet_grid(~potus.buckets)
###
```
I don't get any additional insights from the above plot.

Now use ggpairs to discover relationships in different periods. ggpairs provide one way to visualize relationships that might have been missed so far.  
```{r ggpairs}
#Use normalized scale
df_sub<- subset(df,select=c(n_sp_500,
n_gdp_growth_rate,
n_housing_start,
n_unempl_rate,
n_q_inf_rate,
potus.buckets))
#Rename the columns in order to make the labels more readable
colnames(df_sub)<-
c('sp500',
'gdp', 
'hous_st', 
'unemploy', 
'inflat', 
'potus')
ggpairs(subset(df_sub, potus %in% "(2009,2015]"),
title = "Under Obama", axisLabels = "none")
ggpairs(subset(df_sub, potus %in% "(2001,2009]"),
title = "Under Bush", axisLabels = "none" )
ggpairs(subset(df_sub, potus %in% "(1993,2001]"),
title = "Under Clinton", axisLabels = "none")
ggpairs(subset(df_sub, potus %in% "(1981,1993]"),
title = "Under Reagan/Bush", axisLabels = "none")
ggpairs(subset(df_sub, potus %in% "(1977,1981]"),
title = "Under Carter", axisLabels = "none")
ggpairs(subset(df_sub, potus %in% "(1969,1977]"),
title = "Under Nixon/Ford", axisLabels = "none")
ggpairs(subset(df_sub, potus %in% "(1966,1969]"),
title = "Under Johnson", axisLabels = "none")
####
```
ggpairs confirm earlier observations that economic data correlates with the SP 500 for short periods of time and for some of the periods.  None has correlation all the time. Unemployment seems to be a "useful" variable in 5 of the 7 periods.  Others come in and out. It is also interesting that in the short-term, the S&P 500 looks random against the GDP growth rate.  

# Final Plots and Summary
```{r final_plot 1}
#Before plotting, normalize the data in order to fit on the same chart.  Melt frame to color code the lines.

pop[c('n_pop','n_sp500','n_gdp')] <-
scale(pop[c('pop','sp500','gdp')],center=T,scale=T)

n_pop <-subset(pop,select=c(year, n_pop,n_sp500,n_gdp))

#Melt data frame and keep year as the x-variable
pop_long <- melt (n_pop,id.vars="year", variable.name="metrics", 
value.name="val")
```
One of the key observations is the high correlation between the population, GDP and the S&P500. So it should be interesting to see the 3 time series together.
```{r final plot 1}
lp<-ggplot(data=pop_long,aes(x=year, y=val, color=metrics))+
geom_line(linetype="solid",size=2.0)
lp +  scale_colour_discrete(name="Variables",
labels = c('Population', "S&P 500", "GDP"))+
ggtitle("S&P 500, GDP and Population") + 
ylab("Normalized Values") + 
xlab("Year")
####
```
The above shows that the S&P is highly correlated to the GDP which in turn can be explained by population growth.  As we saw earlier, only the GDP can explain S&P growth over a long period of time.  Other economic variables are correlated with the S&P only over a shorter period of time and only for some of these periods.

Unemployment appeared in earlier plots to have an impact on the S&P in several periods.  Let me see that more clearly by plotting unemployment against the S&P and with Housing Starts coded as circle size, all clustered by colored periods.
```{r final plot 2 }
sp<-ggplot(data=df, aes(x=unempl_rate, y=sp_500, color = potus.buckets,
type="div"))+
geom_point(aes(size=n_housing_start+2))+
scale_y_continuous(trans=log10_trans())
sp + scale_colour_discrete(name="Administration",
labels = c('Johnson', 
'Nixon/Ford', 
'Carter', 
'Reagan/Bush',
'Clinton',
'Bush',
'Obama')) +
scale_size_continuous(name="Normalized Housing Start",
labels = c('1 low','2','3','4', '5 high'))+
theme(legend.title = element_text(size=10))+          
ggtitle('S&P500 Vs Unemployment &\nHousing Start') +
ylab('S&P 500') + xlab('Unemployment Rate')
####
```
The negative corrlelation between unemployment and the S&P 500 is clearly visible in the last 25 years.  Unemployment has an unmistakable relationship with the S&P in clusters.  Even as the overall valuation increases over a long period of time, the stock market is negatively impacted by higher unemployment within a cluster.  Housing starts doesn't seem to be a factor.

I want to show how inflation affects the S&P.  It is a scatter plot of the S&P and the GDP, with the circles size split up by the level of inflation and all color coded by periods.  
```{r final plot 3}
fsp<-ggplot(data=df, aes(x=gdp_growth_rate, y=sp_500, 
                         color=potus.buckets,
                         type='div'))+
  geom_point(aes(size=n_q_inf_rate+5))+
  scale_y_continuous(trans=log10_trans())+
  geom_hline(yintercept=150, color="red") 
fsp+scale_colour_discrete(name="Administration",
                          labels = c('Johnson', 
                                     'Nixon/Ford', 
                                     'Carter', 
                                     'Reagan/Bush',
                                     'Clinton',
                                     'Bush',
                                     'Obama')) +
  scale_size_continuous(name="Normalized Quarterly\nInflation Rate",
                        labels = c('0 low','1 low','2','3', '4', '5 high')) +
  theme(legend.title = element_text(size=10))+
  ggtitle('The Effect of Inflation?') +
  ylab('S&P 500') + xlab("GDP Growth Rate")
#####
```
The above plot raises the question: How is inflation related to the SP500.  What stands out is the faster rise of the S&P post versus pre-Reagan.  It grew at a trajectory different from before.  The quarterly gdp growth rate is less scattered post Reagan.  The population growth rate hasn't changed that much throughout these 47 years, yet the S&P grew much faster.  Something seems to have changed at this transition.  What seems to be consistent is the lower inflation during the last 25 years, compared to higher inflation in the earlier 17 year period.  It would interesting to explore this further.  
```{r Reflection}
```
# Reflection
Do economic factors impact the S&P500?  The NY Stock Exchange thinks so (see NYSE pdf in Reference).  But this analysis shows a more nuanced picture.  Factors that can explain SP500 quarterly changes are elusive. I struggled through the 2 variable analysis and couldn't find any clues.  The only unmistakable trend is the increase in population which intuitively contributes to gdp growth.  In turn, that is highly correlated to the S&P500 over time.  The S&P seems to have discernable relationships with economic variables in clusters.  The unmeployment rate, for instance, appears to be useful some of the time, but not always.  Furthermore, a transition seemed to have occurred in the Reagon years when the slope of the S&P curve changed.  Some system-wide changes that were favorable to stock price appreciation appeared to have taken hold.  This transition cannot be explained by population or gdp growth.  Inflation rates were on average higher before the transition. It would be interesting to explore this transition as a follow-on project.


